String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-04-26 2023-1-19/sections/cefa.tex
cefa

In this section, we recall the definition of CEFA and Parikh image.

%\subsection{Cost-Enriched Finite Automaton} \label{subsec:cefa}
The definition of CEFA in this paper is lightly different from the definition in \cite{atva2020}. In \cite{atva2020}, the cost function is defined as a function $\eta: \Sigma \rightarrow \mathbb{N}$. In this paper, we define the cost function as an integer vector whose elements are the incremental value of registers. Furthermore, we add a new linear integer arithmetic constraint $\theta$ to each final state of the CEFA, which restricts the value of registers. Two types of definitions have the same expressive ability on the $SAT_{CL}$ problem.
\begin{definition}[Cost-Enriched Finite Automaton]
  A cost-enriched finite automaton $\aut$ is a 7-tuple $(Q, \Sigma, \delta, q_I, F, R, \theta)$ where
  \begin{itemize}
    \item $Q,\Sigma,q_I,F$ is defined as NFA,
    \item $R = (r_1\cdots r_n)$ is a vector of mutually distinct cost registers,
    \item $\delta$ is a transition set whose elements are tuples $(q, c, q', \myvec{v})$ where $q, q'$ are states of $Q$, $c$ is a letter in alphabet $\Sigma\cup\{\epsilon\}$ and $\myvec{v}$ is the cost update function for registers, which is an integer vector whose $i$-th element is the incremental value of register $r_i$.  $(q, a, q', \myvec{v})$ is written as $q\xrightarrow[\myvec{v}]{a} q'$ for readability.
    \item $\theta: F\rightarrow \varphi$ is a linear integer arithmetic constraint function on final states. $\theta$ is called \emph{accepting condition}.
  \end{itemize}
  The value of each register $r_i$ in $R$ is written as $\mathcal{V}(r_i)$ and is initiated to 0 at the initial state. A \emph{run} of $\aut$ on string $a_1\cdots a_m$ is a transition sequence $q_I\xrightarrow[\myvec{v_1}]{a_1}q_1\cdots q_{m-1}\xrightarrow[\myvec{v_m}]{a_m}q_m$  and $\mathcal{V}(r_i) = \displaystyle\sum_{j=1}^m \myvec{v_j}[i], i\in [1,n]$ is the value of $r_i$ after the run. $\theta[R/\mathcal{V}(R)](q_m)$ is obtained from $\theta(q_m)$ by replacing each register $r_i$ to its value $\mathcal{V}(r_i)$. The run is \emph{accepting} if $q_m\in F$ and $\theta[R/\mathcal{V}(R)](q_m)$ is satisfiable. $\top$ is the valid formula that is always satisfiable. A string $w$ is accepted by $\aut$ if it has an accepting run of $\aut$. The language of $\aut$ is the set of strings accepted by $\aut$, denoted by $\lan(\aut)$.

>>>>

In this section, we define the model of cost-enriched finite state automata (CEFA), which was introduced in \cite{atva2020} and will be used to solve the satisfiability problem of RECL later on.

%\subsection{Cost-Enriched Finite Automaton} \label{subsec:cefa}
The definition of CEFA below is slightly different from the definition in~\cite{atva2020} in the sense that accepting conditions specified by LIA formulas, where the free variables are the registers, are attached to final states. 
%
%In~\cite{atva2020}, the cost function is defined as a function $\eta: \Sigma \rightarrow \mathbb{N}$. In this paper, we define the cost function as an integer vector whose elements are the incremental value of registers. Furthermore, we add a new linear integer arithmetic constraint $\theta$ to each final state of the CEFA, which restricts the value of registers. Two types of definitions have the same expressive ability on the $SAT_{CL}$ problem.
\begin{definition}[Cost-Enriched Finite Automaton]
  A cost-enriched finite automaton $\aut$ is a tuple $(R, Q, \Sigma, \delta, q_0, F, \theta)$ where
  \begin{itemize}
  \item $R = \{r_1, \cdots, r_n\}$ is a finite set of registers, 
    \item $Q, \Sigma, q_0, F$ are as in the definition of NFA,
%    \item $R = (r_1\cdots r_n)$ is a vector of mutually distinct cost registers,
    \item $\delta \subseteq Q \times \Sigma_\varepsilon \times Q \times (\Int^n)^R$ is a transition set whose elements are tuples $(q, c, q', \myvec{v})$ where $q, q'$ are states of $Q$, $c$ is a letter in alphabet $\Sigma\cup\{\epsilon\}$ and $\myvec{v}$ is the cost update function for registers, which is an integer vector whose $i$-th element is the incremental value of register $r_i$.  $(q, a, q', \myvec{v})$ is written as $q\xrightarrow[\myvec{v}]{a} q'$ for readability.
    \item $\theta: F\rightarrow \varphi$ is a linear integer arithmetic constraint function on final states. $\theta$ is called \emph{accepting condition}.
  \end{itemize}
  The value of each register $r_i$ in $R$ is written as $\mathcal{V}(r_i)$ and is initiated to 0 at the initial state. A \emph{run} of $\aut$ on string $a_1\cdots a_m$ is a transition sequence $q_I\xrightarrow[\myvec{v_1}]{a_1}q_1\cdots q_{m-1}\xrightarrow[\myvec{v_m}]{a_m}q_m$  and $\mathcal{V}(r_i) = \displaystyle\sum_{j=1}^m \myvec{v_j}[i], i\in [1,n]$ is the value of $r_i$ after the run. $\theta[R/\mathcal{V}(R)](q_m)$ is obtained from $\theta(q_m)$ by replacing each register $r_i$ to its value $\mathcal{V}(r_i)$. The run is \emph{accepting} if $q_m\in F$ and $\theta[R/\mathcal{V}(R)](q_m)$ is satisfiable. $\top$ is the valid formula that is always satisfiable. A string $w$ is accepted by $\aut$ if it has an accepting run of $\aut$. The language of $\aut$ is the set of strings accepted by $\aut$, denoted by $\lan(\aut)$.

### 2023-04-26 2023-1-19/sections/introduction.tex
cefa

\paragraph*{Contribution} In this work, we focus on the class of string constraints with regular membership and string-length function, where the counting operators may occur in regular expressions (called RECL for brevity). We propose an automata-theoretical approach for solving RECL constraints. Our main idea is to represent the counting operators by registers in automata, instead of unfolding them explicitly.  Our approach is built upon an automata model called cost-enriched finite automata (CEFA) that was proposed in~\cite{atva2020}. Moreover, the string-length function can also be modeled by registers in CEFA. Therefore, the satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA.

@@

>>>>

\paragraph*{Contribution} In this work, we focus on the class of string constraints with regular membership and string-length function, where the counting operators may occur in regular expressions (called RECL for brevity). We propose an automata-theoretical approach for solving RECL constraints. Our main idea is to represent the counting operators by registers in automata, instead of unfolding them explicitly.  Our approach is built upon an automata model called cost-enriched finite automata (CEFA), which was proposed in~\cite{atva2020} to solve string constraints involving the integer data type. Moreover, the string-length function can also be modeled by registers in CEFA. Therefore, the satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA.

@@

At last, we would like to clarify the relationship of this work with~\cite{atva2020}, where CEFA that was introduced. This work differs from \cite{atva2020} in the following aspects: In \cite{atva2020}, the regex-counting was unfolded explicitly. Moreover, in~\cite{atva2020}, the nonemptiness of CEFA was solved by utilizing the model checker nuXmv~\cite{nuxmv}, while in this work, it is solved by integrating the techniques of Parikh image computation, automata-size reduction, under approximation, and SMT solvers for LIA. The experimental results show that our approach achieves better performance than that in~\cite{atva2020}. \zhilin{this point should be supposed by experimental results.}

### 2023-04-26 2023-1-19/sections/preliminaries.tex
cefa

\paragraph*{Strings and languages.}
A string over $\Sigma$ is a (possibly empty) sequence of elements from $\Sigma$,
denoted by $u, v, w, \ldots$. An empty string is denoted by $\varepsilon$.  We write $\Sigma^*$ (resp., $\Sigma^+$) for the set of all (resp. nonempty) strings over $\Sigma$.
%Let 
For a string $u$, we use $|u|$ to denote the number of letters in $u$. In particular, $|\varepsilon|=0$. 
%Moreover, for $a \in \Sigma$, let $|u|_a$ denote the number of occurrences of $a$ in $u$. 
Assume $u=a_0\cdots a_{n-1}$ is nonempty and $i<j \in [0,n-1]$. %Then a \emph{position} of $u$ is a number $i \in [|u|]$ (Note that the first position is $1$, instead of  0). In addition, 
We let $u[i]$ denote $a_i$ and $u[i,j]$ for the substring %of $u$ starting from $i$ and ending with $j$ (i.e., 
$a_i\cdots a_j$. 
%\tl{check later for consistency}\zhilin{the indices start from 0, to be consistent with the semantics of substring and indexof}\zhilin{i do use it in section 4}
%
Let $u, v$ be two strings. We use $u \cdot v$ to denote the \emph{concatenation} of $u$ and $v$. A language $L$ over $\Sigma$ is a subset of strings.  
Let $L_1, L_2$ be two languages. Then the concatenation of $L_1$ and $L_2$, denoted by $L_1 \cdot L_2$,  is defined as $\{u \cdot v \mid u \in L_1, v \in L_2\}$. The union (resp. intersection) of $L_1$ and $L_2$, denoted by $L_1 \cup L_2$  (resp. $L_1 \cap L_2$), is defined as $\{u \mid u \in L_1 \mbox{ or } u \in L_2\}$ (resp. $\{u \mid u \in L_1 \mbox{ and } u \in L_2\}$). The complement of $L_1$, denoted by $\overline{L_1}$, is defined as $\{u \mid u \in \Sigma^*, u \not \in L_1\}$. 
The difference of $L_1$ and $L_2$, denoted by $L_1 \setminus L_2$, is defined as $L_1 \cap \overline{L_2}$. 
For a language $L$ and $n \in \Nat$, we define $L^n$ inductively as follows: $L^0= \{\varepsilon\}$, $L^{n+1} = L \cdot L^n$ for every $n \in \Nat$. 
Finally, define $L^* = \bigcup \limits_{n \in \Nat} L^n$.
%, that is, the string $w$ such that $|w|= |u| + |v|$ and for each $i \in [0, |u|-1]$, $w[i]= u[i]$, and for each $i \in [0,|v|-1]$, $w[|u|+i]=v[i]$.

\paragraph*{Finite state automata.} 
A \emph{(nondeterministic) finite state automaton} (NFA)  is a tuple $\NFA=(Q, \Sigma, \delta, I, F)$, where $Q$ is a finite set of states, $\Sigma$ is a finite alphabet, $\delta \subseteq Q \times \Sigma \times Q$ is the transition relation, $I,F \subseteq Q$ are the set of initial and final states respectively. For readability, we write a transition $(q, a, q') \in \delta$ as $q \xrightarrow[\delta]{a} q'$ (or simply $q \xrightarrow{a} q'$). %Moreover, when $\delta$ is clear from context, we omit $\delta$ in $q \xrightarrow[\delta]{a} q'$ and write $q \xrightarrow{a} q'$. 
The \emph{size} of an NFA $\NFA$, denoted by $|\NFA|$, is defined as the number of transitions of $\NFA$.
%
A \emph{run} of $\NFA$ on a string $w = a_1 \cdots a_n$ is a sequence of transitions $q_0 \xrightarrow{a_1} q_1 \cdots q_{n-1} \xrightarrow{a_n} q_n$ with $q_0 \in I$. The run is \emph{accepting} if $q_n \in F$.
A string $w$ is accepted by an NFA $\NFA$ if there is an accepting run of $\NFA$ on $w$. In particular, the empty string $\varepsilon$ is accepted by $\NFA$ if $I \cap F \neq \emptyset$. The language of $\NFA$, denoted by $\Lang(\NFA)$, is the set of strings accepted by $\NFA$. 
%
An NFA $\NFA$ is said to be \emph{deterministic} if $I$ is a singleton and, for every $q \in Q$ and $a \in \Sigma$, there is at most one state $q' \in Q$ such that $(q, a, q') \in \delta$.
%
It is well-known that finite automata capture regular languages precisely. Moreover, the class of regular languages is closed under union, intersection, Kleene star, complement, and language difference \cite{HU79}.

>>>>

\paragraph*{Strings and languages.}
A string over $\Sigma$ is a (possibly empty) sequence of elements from $\Sigma$,
denoted by $u, v, w, \ldots$. An empty string is denoted by $\varepsilon$. We use $\Sigma_\varepsilon$ to denote $\Sigma \cup \{\varepsilon\}$. We write $\Sigma^*$ (resp., $\Sigma^+$) for the set of all (resp. nonempty) strings over $\Sigma$.
%Let 
For a string $u$, we use $|u|$ to denote the number of letters in $u$. In particular, $|\varepsilon|=0$. 
%Moreover, for $a \in \Sigma$, let $|u|_a$ denote the number of occurrences of $a$ in $u$. 
Assume $u=a_0\cdots a_{n-1}$ is nonempty and $i<j \in [0,n-1]$. %Then a \emph{position} of $u$ is a number $i \in [|u|]$ (Note that the first position is $1$, instead of  0). In addition, 
We let $u[i]$ denote $a_i$ and $u[i,j]$ for the substring %of $u$ starting from $i$ and ending with $j$ (i.e., 
$a_i\cdots a_j$. 
%\tl{check later for consistency}\zhilin{the indices start from 0, to be consistent with the semantics of substring and indexof}\zhilin{i do use it in section 4}
%
Let $u, v$ be two strings. We use $u \cdot v$ to denote the \emph{concatenation} of $u$ and $v$. A language $L$ over $\Sigma$ is a subset of strings.  
Let $L_1, L_2$ be two languages. Then the concatenation of $L_1$ and $L_2$, denoted by $L_1 \cdot L_2$,  is defined as $\{u \cdot v \mid u \in L_1, v \in L_2\}$. The union (resp. intersection) of $L_1$ and $L_2$, denoted by $L_1 \cup L_2$  (resp. $L_1 \cap L_2$), is defined as $\{u \mid u \in L_1 \mbox{ or } u \in L_2\}$ (resp. $\{u \mid u \in L_1 \mbox{ and } u \in L_2\}$). The complement of $L_1$, denoted by $\overline{L_1}$, is defined as $\{u \mid u \in \Sigma^*, u \not \in L_1\}$. 
The difference of $L_1$ and $L_2$, denoted by $L_1 \setminus L_2$, is defined as $L_1 \cap \overline{L_2}$. 
For a language $L$ and $n \in \Nat$, we define $L^n$ inductively as follows: $L^0= \{\varepsilon\}$, $L^{n+1} = L \cdot L^n$ for every $n \in \Nat$. 
Finally, define $L^* = \bigcup \limits_{n \in \Nat} L^n$.
%, that is, the string $w$ such that $|w|= |u| + |v|$ and for each $i \in [0, |u|-1]$, $w[i]= u[i]$, and for each $i \in [0,|v|-1]$, $w[|u|+i]=v[i]$.

\paragraph*{Finite state automata.} 
A \emph{(nondeterministic) finite state automaton} (NFA)  is a tuple $\NFA=(Q, \Sigma, \delta, q_0, F)$, where $Q$ is a finite set of states, $\Sigma$ is a finite alphabet, $\delta \subseteq Q \times \Sigma \times Q$ is the transition relation, $q_0 \in Q$ is the initial state, and $F \subseteq Q$ is the set of final states. For readability, we write a transition $(q, a, q') \in \delta$ as $q \xrightarrow[\delta]{a} q'$ (or simply $q \xrightarrow{a} q'$). %Moreover, when $\delta$ is clear from context, we omit $\delta$ in $q \xrightarrow[\delta]{a} q'$ and write $q \xrightarrow{a} q'$. 
The \emph{size} of an NFA $\NFA$, denoted by $|\NFA|$, is defined as the number of transitions of $\NFA$.
%
A \emph{run} of $\NFA$ on a string $w = a_1 \cdots a_n$ is a sequence of transitions $q_0 \xrightarrow{a_1} q_1 \cdots q_{n-1} \xrightarrow{a_n} q_n$. The run is \emph{accepting} if $q_n \in F$.
A string $w$ is accepted by an NFA $\NFA$ if there is an accepting run of $\NFA$ on $w$. In particular, the empty string $\varepsilon$ is accepted by $\NFA$ if $I \cap F \neq \emptyset$. The language of $\NFA$, denoted by $\Lang(\NFA)$, is the set of strings accepted by $\NFA$. 
%
An NFA $\NFA$ is said to be \emph{deterministic} if $I$ is a singleton and, for every $q \in Q$ and $a \in \Sigma$, there is at most one state $q' \in Q$ such that $(q, a, q') \in \delta$.
%
It is well-known that finite automata capture regular languages precisely. Moreover, the class of regular languages is closed under union, intersection, Kleene star, complement, and language difference \cite{HU79}.
